---
title: 'Exercises for Monte Carlo simulation workshop'
author: "Tamas Nagy"
date: "6/9/2021"
output: 
  html_document:
   theme: spacelab
   code_download: true
   toc: true
   toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE)
```

# How to use this exercise document?

Just write the solution into the chunk and run the whole chunk. To see if you did right, check the solutions in the `solutions` folder!

# Exercise 1: Simulating variables

This exercise document is about generating random variables. Most chuncks have a `set.seed()` function, that sets the random seed, so the results are the same for everyone. In real life, you should set the seed for reproducibility.


## Exercise 1.1: Draw the winning lottery numbers 
Where we draw 5 from 90 unique numbers. Use the `sample()` function.

```{r}
set.seed(123)

sample(1:90, 5)
```

## Exercise 1.2: Simulate 100 (6-sided) loaded (cheater) dice throws
Where the probability to cast one is 5%, but to throw six is 28.33%. Use the `sample()` function.
*Optional*: plot the results (preferably using ggplot2). 

```{r}
set.seed(123)

loaded_rolls <- sample(1:6, 100, replace = TRUE, prob = c(.05, 1/6, 1/6, 1/6, 1/6, .2833))

# show results
loaded_rolls

# plot results
ggplot() + 
    aes(x = loaded_rolls) +
    geom_histogram()

```


## Exercise 1.3: Simulate 10 coin tosses
This time, use the `rbinom()` function.

```{r}
set.seed(123)

rbinom(n = 10, size = 1, prob = .5)
```

## Exercise 1.4: Simulate reaction time values 
For Group A (drunk) and Group B (sober). Each group has 100 members. The sober group has an average RT of 700, with an SD of 150. The drunk group should have 50ms longer RTs, and the same SD as the sober group. 
Investigate if there is a significant difference between the two groups in average RT?

```{r}
set.seed(123)

group_a <- rnorm(n = 100, mean = 750, sd = 150)
group_b <- rnorm(n = 100, mean = 700, sd = 150)

t.test(group_a, group_b)

```

## Exercise 1.5: Generate 1000 censored normal values, where 
Sometimes you need 
```{r}
set.seed(123)

library(faux)
library(TruncatedNormal)

censored <- rtnorm(n = 1000, mu = 5, sd = 6, lb = 0, ub = 15)



qplot(censored)

mean(censored)
sd(censored)
```

## Exercise 2: Generate a parameter matrix, where we want 
- group sizes from 100 to 1000 replications by increments of 100;
- effect size of .3, .5, .7
- 500 replications of each dataset


```{r}

crossing(group_size = seq(100, 1000, 100),
         effect_size = c(.3, .5, .7),
         )
```


## Exercise 3: Interpolate values for 


